---
title: "An October vaccine signal is possible but not certain"
author: "James C. Capretta and Scott Ganz"
date: "October 2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(mosaic)
options(warn = -1)
set.seed(123456)
```

## Introduction

This replication file replicates the calculations in the paper **An October vaccine signal is possible but not certain**. The code was originally run on R.4.0.1. Updated versions of R may have been released since this replication file was written.

You will need to install the following packages into R to run this code:

* dplyr
* tidyr
* mosaic

The following function **ve_hat_eval** must also be defined. This function takes in a true vaccine efficacy rate (*ve*) and simulates what the observed vaccine efficacy rate would be after various numbers of endpoints.

```{r}
ve_hat_eval = function(ve) {
  rr <- 1 - ve
  x <- rbinom(164, 1, rr/(1 + rr))
  ve_hat <- (1 - cumsum(x == 1)/cumsum(x == 0))[c(32, 50, 53, 62, 75, 92, 100, 106, 120, 150, 151, 164)]
  return(ve_hat)
}
```


## Vaccine Efficacy Bounds

Based on the clinical protocol documents of vaccines currently in Phase III trials, the following vaccine efficacy bounds are defined at a set number of endpoints:

* [WHO Protocol](https://www.who.int/publications/i/item/an-international-randomised-trial-of-candidate-vaccines-against-covid-19)
* [Pfizer Protocol](https://pfe-pfizercom-d8-prod.s3.amazonaws.com/2020-09/C4591001_Clinical_Protocol_0.pdf)

```{r veBounds}
# WHO CRITERIA
who.lwr50 <- -0.14
who.upr50 <- 0.76
who.lwr100 <- 0.23
who.upr100 <- 0.59
who.bound150 <-0.5 
# PFIZER CRITERIA
pfi.lwr32 <- 1-(15/17)
pfi.upr32 <- 1-(6/26)
pfi.lwr62 <- 1-(26/36)
pfi.upr62 <- 1-(15/47)
pfi.lwr92 <- 1-(35/57)
pfi.upr92 <- 1-(25/67)
pfi.upr120 <- 1-(35/85)
pfi.bound164 <-1-(53/111) 
```

## Probability of Success Calculations

### Difference between Pfizer and WHO criteria

"If the vaccine’s true efficacy in the entire population were 55 percent, the probability of a successful trial under Pfizer-BioNTech’s protocol is 9 percentage points lower than the probability of success under the WHO’s recommended criteria."

```{r successProb}
# VE = 55 percent
x <- replicate(10^5, ve_hat_eval(0.55))

# WHO
x.res <- apply(t(x), 1, function(ve) {
    WHOendpoint1 <- ve[2] >= who.upr50 
    WHOendpoint2 <- ve[7] >= who.upr100 & ve[2] > who.lwr50
    WHOendpoint3 <- ve[10] >= who.bound150 & ve[2] > who.lwr50 & ve[7] > who.lwr100
    YES <- ifelse(WHOendpoint1 !=0 | WHOendpoint2 !=0 | WHOendpoint3 !=0, 1, 0)
    return(mean(YES))
  })
  who.success_prob <- mean(x.res)
  
# Pfizer
x.res <- apply(t(x), 1, function(ve) {
    PFIendpoint1 <- ve[1] >= pfi.upr32
    PFIendpoint2 <- ve[4] >= pfi.upr62 & ve[1] > pfi.lwr32
    PFIendpoint3 <- ve[6] >= pfi.upr92 & ve[1] > pfi.lwr32 & ve[4] > pfi.lwr62
    PFIendpoint4 <- ve[9] >= pfi.upr120 & ve[1] > pfi.lwr32 & ve[4] > pfi.lwr62 & ve[6] > pfi.lwr92
    PFIendpoint5 <- ve[12] >= pfi.bound164 & ve[1] > pfi.lwr32 & ve[4] > pfi.lwr62 & ve[6] > pfi.lwr92
    YES <- ifelse(PFIendpoint1 !=0 | PFIendpoint2 !=0 | PFIendpoint3 !=0 | PFIendpoint4 !=0 | PFIendpoint5 !=0, 1, 0)
    return(mean(YES))
  })
  pfi.success_prob <- mean(x.res)
  
# Difference in probability of success between Pfizer and WHO
difference <- pfi.success_prob - who.success_prob
difference
```

### Likelihood of 32 cases after 2 months

"With 44,000 trial participants (when fully enrolled), 32 cases of COVID-19 disease would almost certainly emerge before two months, if Pfizer-BioNTech’s assumptions about true vaccine efficacy and the rate at which nonvaccinated trial participants contract the disease are correct."

*Pfizer assume a yearly attack rate (ar) of 1.3% (3-month attack rate of 0.325%). See [Pfizer Protocol](https://pfe-pfizercom-d8-prod.s3.amazonaws.com/2020-09/C4591001_Clinical_Protocol_0.pdf) page 13 for more. In the example below, we assume a true vaccine efficacy rate (ve) of 0.6.* 

```{r waitingTime}
sim_waiting_time = function(ss, ar, n, ve, nsim) {
  (sapply(1:nsim, function(x) {
    X <- sort(c(rexp(ss/2, ar/3), rexp(ss/2, (ar/3) * (1 - ve))))
    X[X > 0.5][n]
  }))
}

x <- sim_waiting_time(44000, 0.00325, 32, 0.6, 10^5)
prob <- mean(x < 2)
prob
```

### Pfizer probability of success after 32 cases

"The odds of a vaccine meeting the Pfizer-BioNTech’s criteria at 32 cases if the actual efficacy of the vaccine is 70 percent is less than 2 in 5.  If the actual efficacy is only 60 percent, the odds drop to 15 in 100."

```{r successProb2}
# VE = 70 percent
x <- replicate(10^5, ve_hat_eval(0.7))
x.res <- apply(t(x), 1, function(ve) {
    PFIendpoint1 <- ve[1] >= pfi.upr32
    YES <- ifelse(PFIendpoint1 !=0, 1, 0)
    return(mean(YES))
  })
  pfi.success_prob <- mean(x.res)
  pfi.success_prob

# VE = 60 percent  
x <- replicate(10^5, ve_hat_eval(0.6))
x.res <- apply(t(x), 1, function(ve) {
    PFIendpoint1 <- ve[1] >= pfi.upr32
    YES <- ifelse(PFIendpoint1 !=0, 1, 0)
    return(mean(YES))
  })
  pfi.success_prob <- mean(x.res)
  pfi.success_prob
```

### Pfizer probability of success after 62 cases and 92 cases

"With more case counts, the odds improve for meeting the efficacy criteria. If the actual efficacy of a vaccine is 70 percent, then the odds of meeting the FDA and WHO criteria at 62 cases is around 2 in 3, and it rises to 9 in 10 at 92 cases."

```{r successProb3}
x <- replicate(10^5, ve_hat_eval(0.7))

# Odds of success after 62 cases
x.res <- apply(t(x), 1, function(ve) {
    PFIendpoint1 <- ve[1] >= pfi.upr32
    PFIendpoint2 <- ve[4] >= pfi.upr62 & ve[1] > pfi.lwr32
    YES <- ifelse(PFIendpoint1 !=0 | PFIendpoint2 !=0, 1, 0)
    return(mean(YES))
  })
  pfi.success_prob_62cases <- mean(x.res)
  pfi.success_prob_62cases

# Odds of success after 92 cases
x.res <- apply(t(x), 1, function(ve) {
    PFIendpoint1 <- ve[1] >= pfi.upr32
    PFIendpoint2 <- ve[4] >= pfi.upr62 & ve[1] > pfi.lwr32
    PFIendpoint3 <- ve[6] >= pfi.upr92 & ve[1] > pfi.lwr32 & ve[4] > pfi.lwr62
    YES <- ifelse(PFIendpoint1 !=0 | PFIendpoint2 !=0 | PFIendpoint3 !=0, 1, 0)
    return(mean(YES))
  })
  pfi.success_prob_92cases <- mean(x.res)
  pfi.success_prob_92cases
```

