---
title: 'Awaiting the signal: Assessing the efficacy of COVID-19 vaccines - Figures and simulations'
author: "James C. Capretta and Scott Ganz"
date: "September 2020"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(ggthemes)
library(stringr)
library(metR)
options(warn = -1)
```

## Introduction

This replication file is designed to replicate the calculations in the paper **Awaiting the signal: Assessing the efficacy of COVID-19 vaccines** and includes all the R code needed. The calculations were completed with the version R.4.0.1. Updated versions of R may have been released since this replication file was written however the code should continue to run in newer versions without major changes.

You will need to install the following packages into R to run this code:

* dplyr
* readr
* ggplot2
* tidyr
* ggthemes
* stringr
* metR

A seed is also set by the authors so that simulated estimates given in the paper remain the same with each run of the following replication file. The seed does not have to be used however if the seed is not used, results may differ slightly from what can be found within the paper. 

```{r seed}
set.seed(123456)
```

## Simulation parameters

First, lets define the four key factors for efficacy signals determined by the authors:
<ol type="1">
  <li>**Sample Size (ss):** The number of participants taking part in a vaccine trial. This covers both the treatment and the control group. More participants in a trial are better than fewer participants, because more participants will have more primary endpoints for a given attack rate. Thus, the trial can conclude sooner, send a clearer signal of vaccine efficacy, or both.</li>
  <li>**Number of Primary Endpoints (n):** The number of infections which can be identified in either the treatment group or the control group. The number of endpoints tunes the power of the test to distinguish between effective and ineffective vaccines. More endpoints mean a clearer signal of vaccine efficacy. Fewer endpoints offer a less clear signal, but permit the trial to conclude sooner.</li>
  <li>**Attack Rate (ar):** The frequency with which non-vaccinated individuals contract the virus. This also impacts how long the trial takes to conclude. Higher attack rates mean that participants contract the virus more quickly, decreasing the time required to observe a given number of primary endpoints.</li>
  <li>**True Vaccine Efficacy Rate (ve):** How effective the vaccine is. The true vaccine efficacy rate interacts with the sample size, number of endpoints, and attack rate to determine how long the trial lasts and the probability that the trial succeeds. Unlike the sample size, which is fixed as a part of the trial protocol, and the attack rate, which can be estimated from existing data on infection rates, the true efficacy rate is unknown to the trial sponsor beforehand. But, beliefs about vaccine efficacy are a major consideration when determining how many primary endpoints are observed before evaluating the vaccine’s effectiveness. Vaccines that are believed to be very effective require fewer primary endpoints to achieve a high level of confidence that the Phase III trial will be successful.</li>
</ol>

*Note: Through the following replication file, (ve_hat) is defined as the observed vaccine efficacy rate and (ve_star) is defined as the target observed vaccine efficacy rate.*

## False positive analysis

The authors claim that a trial with 30 endpoints and a true vaccine efficacy rate of 0 is expected to have an observed vaccine efficacy rate of 0.5 five percent of the time. Example 1 in the following section of code demonstrates that. Set the true vaccine efficacy rate *ve* to 0, the number of endpoints *n* to 30, and the observed vaccine efficacy rate *ve_hat* to 0.5. The function **pbinom** calculates the probability of observing the observed vaccine efficacy rate of 0.5 if the true vaccine efficacy rate is 0. (note: $rr\_hat/(1+rr\_hat)*n = 10$) The probability of observing this vaccine efficacy rate is 0.0494 $\approx$ 5 percent. However, if the number of endpoints *n* is increased to 300 as in Example 2, the probability of observing the observed vaccine efficacy rate of 0.5 if the true vaccine efficacy rate is 0 reduces to a very small amount, 4.01e-09.

```{r falsepositive}
# 1) Pr(ve_hat = 0.5) if ve = 0 and n = 30
ve <- 0
rr <- 1 - ve
n <- 30
ve_hat <- 0.5
rr_hat <- 1 - ve_hat
pbinom(rr_hat/(1 + rr_hat) * n, n, rr/(1 + rr))

# 2) Pr(ve_hat = 0.5) if ve = 0 and n = 300
n <- 300
pbinom(rr_hat/(1 + rr_hat) * n, n, rr/(1 + rr))
```
## Confidence intervals for observed vaccine efficacy rates

### Required vaccine efficacy rate to satisfy criteria

Next, the authors state that, should trials reach at least 150 confirmed cases and the observed vaccine efficacy rate is greater than 50 percent, then the results of the trial imply that the true efficacy rate falls above 30 percent more than 97.5 times out of 100. To show this, set the observed vaccine efficacy rate *ve_hat* to 0.5 and number of endpoints *n* to 150. Then, using the **qbinom** function, the code stores a value in *x* that represents the 97.5 percentile value for the number of positive cases we would expect in the vaccinated group given *ve_hat*=0.5 and *n*=150. This number is 61 (representing 61 of the 150 confirmed cases are in the vaccinated group). *ve_025* then stores the calculated vaccine efficacy rate if we were to observe 61 of the 150 positive cases in the vaccinated group. It can be seen that *ve_025* is greater than 0.3 (*ve_025* $\approx$ 0.315). Thus, with 97.5 percent confidence, it can be ensured that the true vaccine efficacy rate is greater than 30 percent.

```{r CIforObservedVEs1}
# n = 150 demonstrating that if ve_hat > 0.5
# both criteria are satisfied
ve_hat <- 0.5
rr_hat <- 1 - ve_hat
n <- 150
x <- qbinom(0.975, n, rr_hat/(1 + rr_hat))
ve_025 <- 1 - x/(n - x)
ve_025 > 0.3
```

The authors also claim that, if the observed vaccine efficacy rate is 0.57 or higher after 75 trials, the results of the trial again imply that the true efficacy rate falls above 30 percent more than 97.5 times out of 100. Create a vector *ve0* of 99 values from 0.01 to 0.99 representing potential observed vaccine efficacy rates. Further, set the number of endpoints *n* to 75. Using the **qbinom** function, the code stores values in the vector *q* that represents the 97.5 percentile value for the number of positive cases we would expect in the vaccinated group given the observed vaccine efficacy rates found in *ve0* and given *n*=75. After calculating the observed vaccine efficacy rates for the number of cases in the treatment group given by the values of *q* and storing those efficacy rates in *ve_025*, the code then finds the smallest observed vaccine efficacy rate in the vector *ve0* that corresponds to a vaccine efficacy rate of greater than 0.3 in *ve_025*. That vaccine efficacy rate is 0.57.

```{r CIforObservedVEs2}
# n = 75 demonstrating that if ve_hat >= 0.57
# both criteria are satisfied
ve0 <- 1:99/100
rr0 <- 1 - ve0
n <- 75
q <- qbinom(0.975, n, rr0/(1 + rr0))
ve_025 <- 1 - q/(n - q)
ve_star75 <- min(ve0[ve_025 > 0.3])
ve_star75
```

The following section of code shows that, at *n*=50, we can be 97.5 percent confident that the true vaccine efficacy rate is greater than 30 percent if we have an observed vaccine efficacy rate of 0.7. However, that is not the case if the observed vaccine efficacy rate is 0.6. The structure of the code is the same as testing *n*=150 and *ve_hat*=0.5 above.

```{r CIforObservedVEs3}
# n = 50, ve_hat = 0.7
# both criteria are satisfied
ve_hat <- 0.7
rr_hat <- 1 - ve_hat
n <- 50
x <- qbinom(0.975, n, rr_hat/(1 + rr_hat))
ve_025 <- 1 - x/(n - x)
ve_025 > 0.3

# n = 50, ve_hat = 0.6
# both criteria are NOT satisfied
ve_hat <- 0.6
rr_hat <- 1 - ve_hat
n <- 50
x <- qbinom(0.975, n, rr_hat/(1 + rr_hat))
ve_025 <- 1 - x/(n - x)
ve_025 > 0.3
```

### Required number of cases needed to satisfy criteria

Another way of considering the same issue is, if we have an observed vaccine efficacy rate *ve_hat* of 0.5, what is the minimum number of cases we need to observe in order to be 97.5 percent confident that the true vaccine efficacy rate is greater than 30 percent. The answer to this is 149 cases and can be shown by running the following section of code. Again, the structure of the code is the same as testing *n*=150 and *ve_hat*=0.5 above.

```{r CIforObservedVEs4}
# n = 148, ve_hat = 0.5
# both criteria are NOT satisfied
ve_hat <- 0.5
rr_hat <- 1 - ve_hat
n <- 148
x <- qbinom(0.975, n, rr_hat/(1 + rr_hat))
ve_025 <- 1 - x/(n - x)
ve_025 > 0.3

# n = 149, ve_hat = 0.5
# both criteria are satisfied
ve_hat <- 0.5
rr_hat <- 1 - ve_hat
n <- 149
x <- qbinom(0.975, n, rr_hat/(1 + rr_hat))
ve_025 <- 1 - x/(n - x)
ve_025 > 0.3
```

## Reproducing the confidence intervals in Figure 3

In Figure 3, the authors provide a table of 95 percent confidence intervals based on case counts and observed vaccine efficacy rates. The following section of code reproduces those confidence intervals. Set up two vectors *n* and *ve_hat* with values representing the case counts (number of endpoints) and observed vaccine efficacy rates respectively that are found in Figure 3. Next, create a matrix *d* with 4 columns; one storing the values of *n*, one storing the values of *ve_hat* and one each for the lower and upper bound of the 95 percent confidence interval that will be produced. There should be 16 rows in the matrix *d* with all possible combinations of *n* and *ve_hat* (4x4). 

```{r figure3_1}
# Set values of n and ve_hat
n <- c(50, 100, 150, 300)
ve_hat <- c(0.4, 0.5, 0.6, 0.7)

# Create matrix d with empty columns lwr and upr
d <- expand.grid(n = n, ve_hat = ve_hat) %>% 
  as_tibble %>%
  mutate(lwr = NA,
         upr = NA)
```

The code then runs through a loop for each row of the matrix *d* calculating first the lower bound and then the upper bound for the true vaccine efficacy rate based on the respective values of *n* and *ve*. It then stores those values in the empty columns of *d*. To calculate the lower and upper bounds for the 95 percent confidence interval, the strucutre of the code is the same as has been used above using the **qbinom** function.

Following this, the matrix *d* should be filled with 16 combinations of upper and lower bounds for the 95 percent confidence intervals of the true vaccine efficacy rate. These confidence intervals are the same as those found in Figure 3.

```{r figure3_2}
# Find CI bounds and store in Lwr and upr
for (i in 1:dim(d)[1]) {
  ve_hat <- d$ve_hat[i]
  rr_hat <- 1 - ve_hat
  n <- d$n[i]
  q1 <- qbinom(0.975, n, rr_hat/(1 + rr_hat))
  d$lwr[i] <- 1 - q1/(n - q1)
  q2 <- qbinom(0.025, n, rr_hat/(1 + rr_hat))
  d$upr[i] <- 1 - q2/(n - q2)
}
```

## Simulating wait times

### Waiting time function

The following function **sim_waiting_time** simulates the probability that the desired endpoint count has been reached after a given period of time. Inputted into **sim_waiting_time** is: the sample size (*ss*), the attack rate (*ar*), the number of endpoints (*n*), the true vaccine efficacy rate (*ve*), and the number of simulations (*nsim*).

The function runs through a loop *nsim* times. Each time it goes through, it creates a vector *X*. The vector *X* has length *ss* and represents the simulated wait time until each of the *ss* individuals become infected. The wait time for an individual to become infected is calculated using the exponential function. For the control group, the exponential function uses a sample size of *ss*/2 (half of the overall sample size), and a rate of *ar*/3 (the attack rate for one month). For the treatment group, the sample size is also *ss*/2 however the rate is *ar*/3 multiplied by *(1-ve)* to account for the vaccine.

*X* is then sorted in order from the shortest to the longest wait times. Observations in the first half of the first month are removed as those testing positive in the first 14 days were likely infected before the trial began.

**sim_waiting_time** finally returns the *n*'th wait time in the ordered vector *X*. This number represents the simulated wait time for the *n*'th person in the sample of *ss* total people to become infected. Thus, the function produces *nsim* simulated wait times for the *n*'th person to become infected.

```{r WaitTimeFunction}
sim_waiting_time = function(ss, ar, n, ve, nsim) {
  (sapply(1:nsim, function(x) {
    X <- sort(c(rexp(ss/2, ar/3), rexp(ss/2, (ar/3) * (1 - ve))))
    X[X > 0.5][n]
  }))
}
```

### Running simulated data

Within the section of code below are some examples of the use of the **sim_waiting_time** function that the authors give in the paper. Based off of 10,000 simulations each time, the code first simulates the average waiting time if there is a sample size *ss* of 30,000, an attack rate *ar* of 0.01, the number of endpoints required *n* is 150, and the true vaccine efficacy rate *ve* is 0.5. It stores these 10,000 simulated wait times in the vector *x*. The code then calculates the proportion of the wait times stored in *x* that are less than 2.5 months. That proportion gives approximately a 1 in 2 chance of the trial period ending before two and a half months as claimed by the authors in the paper. Example 1 also outputs the result that there is over a 99 percent chance that the trial period will end before three months with the same parameters.

```{r WaitTimeSimulations1}
# 1) 30k sample, 50 percent ve, 150 cases
x <- sim_waiting_time(30000, 0.01, 150, 0.5, 10^4)
mean(x < 2.5)
mean(x < 3)
```

The same code is run for different combinations of parameters. For example, if the number of endpoints *n* is changed to 100 rather than 150 such as in example 2, there is basically a 100 percent chance that the endpoint count is reached after 2.5 months. The same is the case within just 2 months if the sample size *ss* is increased from 30,000 to 45,000 such as in example 3. The fourth example shows that there is approximately a 1 in 5 chance the endpoint count of 150 is reached if the true vaccine efficacy rate *ve* is 0.6 rather than 0.5. That reduces to a 1 in 20 chance if the true vaccine efficacy rate *ve* is increased to 0.7 as demonstrated in the fifth example. Each of these examples are given in the paper.

```{r WaitTimeSimulations2}
# 2) 30k sample, 50 percent ve, 100 cases
x <- sim_waiting_time(30000, 0.01, 100, 0.5, 10^4)
mean(x < 2.5)

# 3) 45k sample, 50 percent ve, 150 cases
x <- sim_waiting_time(30000*1.5, 0.01, 100, 0.5, 10^4)
mean(x < 2)

# 4) 30k sample, 60 percent ve, 150 cases
x <- sim_waiting_time(30000, 0.01, 150, 0.6, 10^4)
mean(x < 2.5)

# 5) 30k sample, 70 percent ve, 150 cases
x <- sim_waiting_time(30000, 0.01, 150, 0.7, 10^4)
mean(x < 2.5)
```

## Statistical power analysis

In the paper, the authors discuss the tradeoff between the power of the trial to identify effective vaccines and the number of observed cases before evaluating the candidate vaccine’s efficacy. In the code below are the calculations that demonstrate the probability of satisfying the statistical criteria given certain parameters. 

### Calculating the probability of success

With a true vaccine efficacy rate *ve* of 0.6, the number of endpoints *n* at 150 and a target observed vaccine efficacy rate *ve_star* of 0.5 the authors use the **pbinom** function to calculate the probability that the target observed vaccine efficacy rate is actually observed. This produces a value of 0.915 and thus the authors claim there is a 9 in 10 chance of satisfying the statistical criteria given those parameters.

```{r PowerAnalysis0}
# 1) 60 percent ve, 150 cases
ve <- 0.6
rr <- 1 - ve
n <- 150
ve_star <- 0.5
rr_star <- 1 - ve_star
pbinom(rr_star/(1 + rr_star) * n, n, rr/(1 + rr))
```

Recall that at 150 endpoints and an observed vaccine efficacy rate of 0.5, the second statistical criterion of a true vaccine efficacy rate exceeding 30 percent 97.5 percent of the time is implied. However, if we reduce the number of endpoints below 150, this second statistical criterion takes precedent and determines a higher required observed vaccine efficacy rate than 0.5. Thus when calculating the probability of satisfying both statistical criteria when the number of endpoints is reduced to 100, the required observed vaccine efficacy rate must first be obtained. In example 2 below, the target observed vaccine efficacy rate *ve_star100* is calculated using the same code structure as in the section titled "Confidence intervals for observed vaccine efficacy rates" earlier. The **pbinom** function is then used to calculate the probability that the target observed vaccine efficacy rate *ve_star100* is actually observed. At 100 endpoints, there is a 74.4% chance of observing the target observed vaccine efficacy rate ($\approx$ 3 in 4 chance). In example 3 where the number of endpoint *n* is reduced to 50, the probability of observing the target observed vaccine efficacy rate is 53.6% and thus much nearer a 1 in 2 chance as the authors claim in the paper.

```{r PowerAnalysis1}
# 2) 60 percent ve, 100 cases
#    find ve_star100
ve0 <- 1:99/100
rr0 <- 1 - ve0
n <- 100
q <- qbinom(0.975, n, rr0/(1 + rr0))
ve0_025 <- 1 - q/(n - q)
ve_star100 <- min(ve0[ve0_025 > 0.3])
#    find probability of success
rr_star100 <- 1 - ve_star100
pbinom(rr_star100/(1 + rr_star100) * n, n, rr/(1 + rr))

# 3) 60 percent ve, 50 cases
#    find ve_star50
ve0 <- 1:99/100
rr0 <- 1 - ve0
n <- 50
q <- qbinom(0.975, n, rr0/(1 + rr0))
ve0_025 <- 1 - q/(n - q)
ve_star50 <- min(ve0[ve0_025 > 0.3])
#    find probability of success
rr_star50 <- 1 - ve_star50
pbinom(rr_star50/(1 + rr_star50) * n, n, rr/(1 + rr))
```

### Calculating parameters needed for 90 percent chance of success

Another way of considering this same tradeoff is by examining what the true efficacy rate of a vaccine must be in order to achieve a 90 percent probability of success at various case counts. In the section of code below are the two examples the authors give in the paper: at 100 cases and 50 cases. In the first example a vector *x* is created of length 99 storing the number of positive cases that can be expected in the vaccinated group at the 90th percentile given that the number of endpoints *n* is 100 for true vaccine efficacy rates ranging from 0.01 to 0.99. The observed vaccine efficacy rates for this 90th percentile value are then calculated and stored in the vector *ve90*. Finally, the code outputs the lowest true vaccine efficacy rate in *ve* that has it's corresponding value in *ve90* greater than *ve_star100*, the target observed vaccine efficacy rate. This value represents the true vaccine efficacy rate required to have a 90 percent probability of success at 100 endpoints. This value is 66 percent. The same is done for 50 endpoints in example 2 where the observed vaccine efficacy rate required increases to 73 percent.

```{r PowerAnalysis2}
# 1) 100 cases, 30 percent threshold, 90 percent success
ve <- 1:99/100
rr <- 1 - ve
n <- 100
x <- qbinom(0.9, n, rr/(1 + rr))
ve90 <- 1 - x/(n - x)
min(ve[ve90 > ve_star100])

# 2) 50 cases, 30 percent threshold, 90 percent success
ve <- 1:99/100
rr <- 1 - ve
n <- 50
x <- qbinom(0.9, n, rr/(1 + rr))
ve90 <- 1 - x/(n - x)
min(ve[ve90 > ve_star50])
```

### Difference between efficacy and probability of passing Phase III

The authors also highlight the important difference between statements related to beliefs about the efficacy of a vaccine and beliefs about the probability of passing a Phase III trial. A claim that a candidate vaccine has a 3 in 4 chance of satisfying the criteria for regulatory approval in a trial with 150 confirmed cases, for example, is consistent with a claim that the vaccine has a 55 percent expected efficacy rate and this is shown in example 1 below. Set the true vaccine efficacy rate *ve* at 0.55, the number of endpoints *n* at 150, and the observed vaccine efficacy rate *ve_hat* at 0.5. Using the function **pbinom**, the first example below shows that there is approximately a 3 in 4 chance of observing an observed vaccine efficacy rate of 0.5 and thus satisfying the statistical criteria.

```{r PowerAnalysis3a}
# 1) 55 percent ve, 150 cases
ve <- 0.55
rr <- 1 - ve
n <- 150
ve_hat <- 0.5
rr_hat <- 1 - ve_hat
pbinom(rr_hat/(1 + rr_hat) * n, n, rr/(1 + rr))
```

In contrast, a claim that a candidate vaccine has an expected efficacy rate of 75 percent is consistent with a belief that the vaccine has less than a 1 in 20,000 chance of failing to satisfy the statistical criteria in a trial with 150 endpoints. This is shown in example 2.  Set the true vaccine efficacy rate *ve* at 0.75, the number of endpoints *n* at 150, and the observed vaccine efficacy rate *ve_hat* at 0.5. The function **pbinom** is used to find the probability that a vaccine efficacy rate of at least 0.5 will be observed so one minus this value is the probability that a vaccine efficacy rate of at least 0.5 will not be observed. As can be seen in example 2, even when we multiply this value by 20,000, it is still less than one. Thus, it can be concluded that a vaccine with a true efficacy rate of 0.75 has less than a 1 in 20,000 chance of failing to satisfy the statistical criteria in a trial with 150 endpoints.

```{r PowerAnalysis3b}
# 2) 75 percent ve, 150 cases
ve <- 0.75
rr <- 1 - ve
n <- 150
ve_star <- 0.5
rr_star <- 1 - ve_star
(1 - pbinom(rr_star/(1 + rr_star) * n, n, rr/(1 + rr))) * 20000
```

## Evaluating at intermediate endpoint counts

The WHO recommends that a trial be evaluated at 50, 100, and 150 endpoints with vaccine efficacy thresholds of 76%, 59%, and 50%, respectively, at each stage. The following section of code will assess, on average, how many endpoints will be required to test a vaccine with a certain true vaccine efficacy rate given these guidelines.

### Cumulatve observed efficacy rate function

The function **ob_eval** found below is used to find the observed vaccine efficacy rate after 50 endpoint and 100 endpoints. Inputted into **ob_eval** is the true vaccine efficacy rate (*ve*). The function produces a vector *x* of length 100 compiled of zeros and ones. These values indicate whether an infected observation is in the treatment group (1) or the control group (0) and is simulated based on the true vaccine efficacy rate *ve*. Then, the vector *ve0* takes the vector *x* and outputs a vector of length 2 that contains the cumulative vaccine efficacy rate after 50 trials and 100 trials (equivalent to 50 endpoints and 100 endpoints).

```{r ob_eval}
ob_eval = function(ve) {
  rr <- 1 - ve
  x <- rbinom(100, 1, rr/(1 + rr))
  ve0 <- (1 - cumsum(x == 1)/cumsum(x == 0))[c(50, 100)]
  return(ve0)
}
```

### Running simulated data

First, define the observed vaccine efficacy thresholds at 50 and 100 endpoints as given by the WHO. The upper bound at 50 endpoints is 0.76 and at 100 endpoints is 0.59. Lower bounds of -0.14 at 50 endpoints and 0.23 at 100 endpoints are also set. At these observed vaccine efficacy rates, we can conclude that the vaccine will not meet the required statistical criteria and thus gives reason to end the trial when we find efficacy rates below these values at the respective endpoints.

```{r Bounds}
lwr50 <- -0.14
upr50 <- 0.76
lwr100 <- 0.23
upr100 <- 0.59
```

The code runs the function **ob_eval** for a true vaccine efficacy rate of 0.55 a total of 10^5 times and stores the results in the 2x(10^5) matrix *x50* where each column represents one simulation of **ob_eval(55)** and the two rows represent the cumulative vaccine efficacy rate after 50 endpoints and 100 endpoints respectively.

The code next takes the transpose of the matrix *x50* (becomes a (10^5)x2 matrix). Then, by row, it assess whether the observed vaccine efficacy rate is greater than 0.76 or less than -0.14 in column 1 (after 50 endpoints) and if the observed vaccine efficacy rate is greater than 0.59  or less than 0.23 in the second column (after 100 endpoints), assigning a 1 if the condition is satisfied and 0 if the condition is not. This is stored in the matrix *x55res*.

Next, the code will go through *x50res* row by row and find the proportion of simulations ending after 50 endpoints storing that value in *p1*, the proportion of simulations ending after 100 endpoints storing that value in *p2*, and the proportion of simulations that will need to take 150 endpoints storing that value in *p3*. Finally, it will compute the sum $p1*50 + p2*100 + p3*150$ and this number (129.59) represents the average number of endpoints that will be required for the WHO to approve a vaccine with a true vaccine efficacy rate of 0.55.

```{rEarlySignalsSims1}
# 1) 55 percent ve
x55 <- replicate(10^5, ob_eval(0.55))
x55res <- apply(t(x55), 1, function(ve) {
  c1 <- ve[1] > upr50 | ve[1] < lwr50
  c2 <- ve[2] > upr100 | ve[2] < lwr100
  return(c(c1, c2))
})
p1 <- mean(x55res[1,] == 1)
p2 <- mean(x55res[2,] == 1 & x55res[1,] == 0)
p3 <- mean(x55res[2,] == 0 & x55res[1,] == 0)
50 * p1 + 100 * p2 + 150 * p3
```

Example 2 does the same for a true vaccine efficacy rate of 0.75 and shows that, on average, it will take 78.27 days for the WHO to be able to approve a vaccine with that true efficacy rate.

```{r EarlySignalsSims2}
# 2) 75 percent ve
x75 <- replicate(10^5, ob_eval(0.75))
x75res <- apply(t(x75), 1, function(ve) {
  c1 <- ve[1] > upr50 | ve[1] < lwr50
  c2 <- ve[2] > upr100 | ve[2] < lwr100
  return(c(c1, c2))
})
p1 <- mean(x75res[1,] == 1)
p2 <- mean(x75res[2,] == 1 & x75res[1,] == 0)
p3 <- mean(x75res[2,] == 0 & x75res[1,] == 0)
50 * p1 + 100 * p2 + 150 * p3
```

### Replicating Figure 5 results

The authors further provide a table of probabilities of success in Figure 5 depending on varying true vaccine efficacy rates and intermediate endpoint counts. To recreate this table, first create a vector *ve* of the four true vaccine efficacy rates given in Figure 5 (0.5, 0.6, 0.7, and 0.8). Then, create a 4x4 matrix *out* such that the first column holds the values of *ve* and  that there are three columns labeled *p50*, *p100*, and *p150* representing 50, 100, and 150 endpoints respectively. The following for loop will then run through row by row calculating the function **ob_eval** for each row's respective true vaccine efficacy rate. The loop will also fill in the remaining values for each row representing the probability of success at each of the respective endpoint counts. To replicate the format of the table in Figure 5, simply take the transpose of *out*.

```{r}
ve <- 5:8/10

out <- tibble(ve = ve) %>%
  mutate(p50 = 0,
         p100 = 0,
         p150 = 0)

for (i in 1:dim(out)[1]) {
  x <- replicate(10^5, ob_eval(out$ve[i]))
  out$p50[i] <- mean(apply(t(x), 1, function(ve) ve[1] >= upr50))
  out$p100[i] <- mean(apply(t(x), 1, function(ve) ve[2] >= upr100 & ve[1] < upr50 & ve[1] > lwr50))
  out$p150[i] <- mean(apply(t(x), 1, function(ve) ve[3] >= 0.5 & ve[2] < upr100 & ve[2] > lwr100 & ve[1] < upr50 & ve[1] > lwr50))
}
```

